---
title: "Problem Set 2: When Data Lies"
author: "Mizuki Yashiro"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
library(tidyverse)
```

```{r}
gay <- read_csv("data/raw/gay.csv")
gay_r <- read_csv("data/raw/gayreshaped.csv")
ccap <- read_csv("data/raw/ccap2012.csv")
```

# Part 1: The Original Study

## Q1.1

Total observations
```{r}
nrow(gay)
```

Unique study values
```{r}
n_distinct(gay$study)
```

List treatment conditions
```{r}
unique(gay$treatment)
```
There are 69,592 total observations in `gay.csv`. The variable `study` takes 2 unique values (Study 1 and Study 2). The five treatment conditions are:
1. No Contact, 2. Recycling Script by Gay Canvasser, 3. Same-Sex Marriage Script by Gay Canvasser, 4. Recycling Script by Straight Canvasser, 5. Same-Sex Marriage Script by Straight Canvasser.

## Q1.2

```{r q1-2}
s1w1 <- gay |>
  filter(study == 1, wave == 1)

s1w1 |>
  group_by(treatment) |>
  summarize(mean_ssm = mean(ssm), .groups = "drop") |>
  arrange(desc(mean_ssm))
```
In Study 1, Wave 1, the treatment group with the highest average support for same-sex marriage is the "Recycling Script by Gay Canvasser" group, with a mean of 3.130975. However, the differences across groups are small, and all means fall between 3.013474 and 3.130975. This suggests that the groups look fairly similar at baseline, which is consistent with successful random assignment.

## Q1.3

```{r q1-3}
s1w1 |>
  group_by(treatment) |>
  summarize(sd_ssm = sd(ssm), .groups = "drop")
```

The standard deviations of `ssm` are very similar across all treatment groups, ranging from 1.666099 to 1.702533. This suggests that the spread of attitudes toward same-sex marriage is comparable across groups at baseline. In a randomized experiment, we expect not only similar means but also similar variability across groups. Similar standard deviations support the idea that the groups were successfully randomized and are comparable before any treatment occurs.

## Q1.4

```{r q1-4}
s1w1_means <- s1w1 |>
  group_by(treatment) |>
  summarize(mean_ssm = mean(ssm), .groups = "drop")

ggplot(s1w1_means, aes(x = reorder(treatment, mean_ssm), y = mean_ssm)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Study 1, Wave 1: Mean Support for Same-Sex Marriage",
    x = "Treatment Group",
    y = "Mean SSM (1–5 scale)"
  )
```

The bar plot shows that the mean levels of support for same-sex marriage are very similar across all treatment groups at baseline. The differences are small (3.013474 and 3.130975), and no group stands out dramatically. This visual similarity reinforces the idea that the random assignment appears successful, as the groups look comparable before any treatment effects could occur.

## Q1.5
```{r q1-5}
comp <- s1w1 |>
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", 
                          "No Contact")) |>
  group_by(treatment) |>
  summarize(
    mean_ssm = mean(ssm),
    sd_ssm = sd(ssm),
    n = n(),
    .groups = "drop"
  )

comp
```

```{r q1-5-std-diff}
m_t <- comp$mean_ssm[comp$treatment == "Same-Sex Marriage Script by Gay Canvasser"]
m_c <- comp$mean_ssm[comp$treatment == "No Contact"]

s_t <- comp$sd_ssm[comp$treatment == "Same-Sex Marriage Script by Gay Canvasser"]
s_c <- comp$sd_ssm[comp$treatment == "No Contact"]

std_diff <- (m_t - m_c) / sqrt((s_t^2 + s_c^2)/2)

std_diff
```

The standardized difference in means between the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” group is -0.01049935. This is far below the 0.25 threshold typically used to assess balance in randomized experiments. This indicates that the two groups are extremely similar at baseline. Because the groups are well balanced before treatment, any differences observed in later waves can more plausibly be attributed to the treatment rather than pre-existing differences.

## Q1.6

```{r q1-6}
s1w2 <- gay |>
  filter(study == 1, wave == 2)

s1w2 |>
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser",
                          "No Contact")) |>
  group_by(treatment) |>
  summarize(mean_ssm = mean(ssm), .groups = "drop")
```


```{r q1-6-diff}
m_t2 <- s1w2 |>
  filter(treatment == "Same-Sex Marriage Script by Gay Canvasser") |>
  summarize(m = mean(ssm)) |>
  pull(m)

m_c2 <- s1w2 |>
  filter(treatment == "No Contact") |>
  summarize(m = mean(ssm)) |>
  pull(m)

diff_wave2 <- m_t2 - m_c2

diff_wave2
```

In Study 1, Wave 2, the mean support for same-sex marriage is approximately 3.14 in the “Same-Sex Marriage Script by Gay Canvasser” group and 3.04 in the “No Contact” group. The difference in means (treated minus control) is 0.09987463 points. Because the two groups were well balanced at baseline, this difference suggests a treatment effect. If the data were valid, we would interpret this as evidence that a single 20-minute conversation with a gay canvasser increased support for same-sex marriage.

# Part 2: The Forensic Discovery

## Q2.1

```{r q2-1}
# Number of observations
nrow(gay_r)
nrow(ccap)

# Column names
colnames(gay_r)
colnames(ccap)
```

The dataset `gayreshaped.csv` contains 11,948 observations and includes the variables study, treatment, therm1, therm2, therm3, and therm4. These thermometer variables represent attitudes toward gay men and lesbians measured across survey waves. The dataset `ccap2012.csv` contains 43,998 observations and includes the variables ...1, caseid, and gaytherm. The variable gaytherm represents feeling thermometer scores from a large national survey.

## Q2.2

```{r q2-2}
# Study 1 baseline
study1_baseline <- gay_r |>
  filter(study == 1) |>
  pull(therm1)

mean(study1_baseline)
median(study1_baseline)
sd(study1_baseline)

# CCAP thermometer (remove NAs)
ccap_therm <- ccap |>
  filter(!is.na(gaytherm)) |>
  pull(gaytherm)

mean(ccap_therm)
median(ccap_therm)
sd(ccap_therm)
```
The Study 1 baseline thermometer has a mean of approximately 58.37625 and a standard deviation of 28.50288. The CCAP national survey thermometer has a mean of approximately 58.70517 and a standard deviation of 29.36892. The summary statistics are strikingly similar across the two datasets. Given that these are supposed to be independent samples collected separately, such close agreement in both mean and standard deviation is suspicious and suggests the two distributions may be unusually similar.

## Q2.3

```{r q2-3}
# Create combined dataset
combined <- tibble(
  value = c(study1_baseline, ccap_therm),
  source = c(
    rep("LaCour Study 1 Baseline", length(study1_baseline)),
    rep("CCAP 2012 National Survey", length(ccap_therm))))

ggplot(combined, aes(x = value)) +
  geom_histogram(breaks = seq(0, 100, by = 5), color = "black") +
  facet_wrap(~ source) +
  labs(
    title = "Comparison of Feeling Thermometer Distributions",
    x = "Feeling Thermometer (0–100)",
    y = "Count")
```

The two histograms look strikingly similar in overall shape. Both distributions show similar clustering patterns across the 0–100 scale, including noticeable mass around common rounded values. Although the CCAP sample is larger (so the counts are higher), the underlying structure of the distributions appears nearly identical. Given that these datasets are supposed to come from independent samples collected separately, such close visual similarity is suspicious. It raises the possibility that the Study 1 baseline data may not have been independently collected.

## Q2.4

```{r q2-4}
# Study 1 control group
s1_control <- gay_r |>
  filter(study == 1, treatment == "No Contact")

cor_s1 <- cor(s1_control$therm1, s1_control$therm2, use = "complete.obs")
sd_change_s1 <- sd(s1_control$therm2 - s1_control$therm1, na.rm = TRUE)

# Study 2 control group
s2_control <- gay_r |>
  filter(study == 2, treatment == "No Contact")

cor_s2 <- cor(s2_control$therm1, s2_control$therm2, use = "complete.obs")
sd_change_s2 <- sd(s2_control$therm2 - s2_control$therm1, na.rm = TRUE)


cor_s1
sd_change_s1

cor_s2
sd_change_s2
```

In Study 1, the correlation between therm1 and therm2 in the control group is approximately  0.9975817, with a standard deviation of within-person change of about 1.987678. This suggests that individuals’ responses barely changed between waves and are almost perfectly predicted by their baseline responses. In contrast, Study 2 shows a correlation of approximately 0.9734449 and a standard deviation of change of about 6.605738. This pattern is consistent with real longitudinal survey data, where responses are highly but not perfectly correlated and individuals exhibit meaningful variation over time. The near-perfect correlation in Study 1 is implausibly high and strongly suggests that the follow-up data may have been artificially constructed from baseline responses.

## Q2.5

```{r q2-5}
library(patchwork)

p1 <- ggplot(s1_control, aes(x = therm1, y = therm2)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Study 1 Control",
       x = "therm1",
       y = "therm2")

p2 <- ggplot(s2_control, aes(x = therm1, y = therm2)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Study 2 Control",
       x = "therm1",
       y = "therm2")

p1 + p2
```

In the Study 1 control group, the scatter plot shows points lying almost perfectly on the 45-degree line, with extremely little variation around it. This suggests that therm2 is nearly identical to therm1 for most respondents. Such near-perfect replication is implausible in real survey data, where measurement error and minor opinion changes should introduce noticeable variation. In contrast, the Study 2 control group shows a clear positive relationship but with substantial scatter around the line. This pattern is consistent with real longitudinal data, where responses are strongly correlated over time but not perfectly stable. Study 2 therefore appears much more realistic.

## Q2.6

The evidence strongly suggests that Study 1’s data were fabricated rather than collected through a real experiment. First, the baseline distribution of feeling thermometer scores in Study 1 is strikingly similar to the CCAP national survey, with nearly identical means (58.38 vs. 58.71) and standard deviations (28.50 vs. 29.37). Given that these datasets were supposedly collected independently, such close agreement in both central tendency and dispersion is statistically unlikely and raises serious concerns about data independence.

Second, the re-test correlation in the Study 1 control group is approximately 0.998, which is implausibly high for real longitudinal survey data. In contrast, Study 2 shows a more realistic correlation of 0.973. Real respondents’ answers typically fluctuate due to measurement error, rounding, and genuine opinion changes.

Third, the standard deviation of within-person change in Study 1 is only about 1.99, compared to 6.61 in Study 2. This suggests that respondents’ follow-up responses in Study 1 barely changed from baseline, which is inconsistent with normal survey variability.

Taken together, these patterns strongly suggest that Study 1’s follow-up data were artificially generated from baseline values with minimal added noise. A fabricator attempting to create convincing data might copy a known national distribution and then add small random perturbations, inadvertently producing implausibly perfect correlations and unrealistically stable responses.

# Part 3: What Real Data Looks Like

## Q3.1

```{r q3-1}
# Filter to Study 2
study2 <- gay_r |>
  filter(study == 2)

# Number of observations
nrow(study2)

# Baseline summary by treatment group
study2 |>
  group_by(treatment) |>
  summarize(
    mean_therm1 = mean(therm1, na.rm = TRUE),
    sd_therm1 = sd(therm1, na.rm = TRUE),
    n = n(),
    .groups = "drop")
```

Study 2 contains 2,441 observations. At baseline, the mean thermometer score in the No Contact group is 57.88612 (SD = 28.69655), while the Same-Sex Marriage Script by Gay Canvasser group has a mean of 59.35057 (SD = 28.41578).

The means are relatively close and the standard deviations are nearly identical, suggesting that the treatment and control groups in Study 2 are reasonably balanced at baseline.

## Q3.2

```{r q3-2}
# Extract baseline therm1 for both studies
baseline_s1 <- gay_r |>
  filter(study == 1) |>
  pull(therm1)

baseline_s2 <- gay_r |>
  filter(study == 2) |>
  pull(therm1)

baseline_combined <- tibble(
  value = c(baseline_s1, baseline_s2),
  study = c(
    rep("Study 1", length(baseline_s1)),
    rep("Study 2", length(baseline_s2))))

ggplot(baseline_combined, aes(x = value)) +
  geom_histogram(breaks = seq(0, 100, by = 5), color = "black") +
  facet_wrap(~ study) +
  labs(
    title = "Baseline Thermometer Distributions: Study 1 vs Study 2",
    x = "Feeling Thermometer (0–100)",
    y = "Count")
```

The baseline distributions in Study 1 and Study 2 differ visually. Study 1 shows sharper spikes at common values (particularly around 50 and 100), producing a more structured and patterned appearance. In contrast, Study 2’s distribution appears more irregular and uneven across bins. The Study 2 distribution looks more like typical survey data, with natural variability and less pronounced clustering. This contrast reinforces the earlier suspicion that Study 1’s data may have been artificially constructed.


## Q3.3

```{r q3-3}
study2_comp <- study2 |>
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser",
                          "No Contact")) |>
  group_by(treatment) |>
  summarize(
    mean_therm1 = mean(therm1, na.rm = TRUE),
    sd_therm1 = sd(therm1, na.rm = TRUE),
    n = n(),
    .groups = "drop")

study2_comp
```
```{r q3-3-std}
m_t_s2 <- study2_comp$mean_therm1[study2_comp$treatment == "Same-Sex Marriage Script by Gay Canvasser"]
m_c_s2 <- study2_comp$mean_therm1[study2_comp$treatment == "No Contact"]

s_t_s2 <- study2_comp$sd_therm1[study2_comp$treatment == "Same-Sex Marriage Script by Gay Canvasser"]
s_c_s2 <- study2_comp$sd_therm1[study2_comp$treatment == "No Contact"]

std_diff_s2 <- (m_t_s2 - m_c_s2) / sqrt((s_t_s2^2 + s_c_s2^2)/2)

std_diff_s2
```

The standardized difference in baseline thermometer scores between the treatment and control groups in Study 2 is 0.05128243. This is well below the 0.25 threshold typically used to assess balance in randomized experiments. This suggests that Study 2 exhibits good baseline balance between groups. Unlike Study 1, however, the balance in Study 2 appears realistic rather than implausibly perfect.

## Q3.4

```{r q3-4}
study2 |>
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser",
                          "No Contact")) |>
  group_by(treatment) |>
  summarize(
    mean_therm2 = mean(therm2, na.rm = TRUE),
    .groups = "drop")
```

In Study 2, the mean post-treatment thermometer score (therm2) is 57.63523 in the No Contact group and 61.20311 in the Same-Sex Marriage Script by Gay Canvasser group. The difference in post-period means (treated minus control) is 3.56788 points. This suggests a positive treatment effect in Study 2, though we must assess the uncertainty around this estimate before drawing strong conclusions.

## Q3.5

```{r q3-5}
# Get variances and sample sizes
study2_var <- study2 |>
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser",
                          "No Contact")) |>
  group_by(treatment) |>
  summarize(
    var_therm2 = var(therm2, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

study2_var
```

```{r q3-5-ci}
var_t <- study2_var$var_therm2[study2_var$treatment == "Same-Sex Marriage Script by Gay Canvasser"]
var_c <- study2_var$var_therm2[study2_var$treatment == "No Contact"]

n_t <- study2_var$n[study2_var$treatment == "Same-Sex Marriage Script by Gay Canvasser"]
n_c <- study2_var$n[study2_var$treatment == "No Contact"]

SE <- sqrt(var_t/n_t + var_c/n_c)

diff_est <- 61.20311 - 57.63523

lower <- diff_est - 1.96 * SE
upper <- diff_est + 1.96 * SE

SE
lower
upper
```

The standard error of the difference in means is 1.160643. The 95% confidence interval for the treatment effect is (1.293019, 5.842741). Because this confidence interval does not include zero, we reject the null hypothesis of no difference between groups. This suggests that the treatment effect in Study 2 is statistically significant at the 0.05 level.

## Q3.6

```{r q3-6}
# Study 2 control
s2_control <- gay_r |>
  filter(study == 2, treatment == "No Contact")

cor_12 <- cor(s2_control$therm1, s2_control$therm2, use = "complete.obs")
cor_13 <- cor(s2_control$therm1, s2_control$therm3, use = "complete.obs")
cor_14 <- cor(s2_control$therm1, s2_control$therm4, use = "complete.obs")

cor_12
cor_13
cor_14
```

In the Study 2 control group, the correlation between therm1 and therm2 is 0.9734449, between therm1 and therm3 is 0.9594085, and between therm1 and therm4 is 0.9709017. The correlations remain high over time, indicating that attitudes are relatively stable. However, they are not perfect and show some variation across waves. This pattern is consistent with real longitudinal survey data, where responses are strongly correlated but subject to measurement error and genuine changes over time.

## Q3.7

Study 2 differs from Study 1 in several important ways. In Study 2, the re-test correlation in the control group is approximately 0.973 rather than 0.998, and the standard deviation of within-person change is about 6.61 rather than 1.99. These values are consistent with realistic longitudinal survey data, where responses are highly but not perfectly stable and individuals exhibit meaningful variation over time.

In contrast, Study 1’s near-perfect re-test correlation and minimal within-person change suggest that follow-up responses were artificially constructed from baseline values. Real treatment effects are typically modest and embedded within substantial individual-level variability, as seen in Study 2’s effect of approximately 3.57 points alongside realistic dispersion and measurement noise.

This case demonstrates that simple measurement tools — including means, standard deviations, histograms, and correlations — are powerful diagnostic instruments. By carefully examining distributional patterns and re-test reliability, researchers were able to detect statistical irregularities that ultimately revealed fabricated data.

# Part 4: Formalizing the Evidence with Regression

## Q4.1

```{r q4-1}
# Study 1 control group
s1_control <- gay_r |>
  filter(study == 1, treatment == "No Contact")

fit_s1 <- lm(therm2 ~ therm1, data = s1_control)

summary(fit_s1)
```

In Study 1, the estimated slope is 0.992454, the R-squared is 0.9952, and the residual standard error is 1.976. This indicates that therm1 almost perfectly predicts therm2, with baseline attitudes explaining more than 99% of the variation in follow-up attitudes. Such an extremely high R-squared and near-unit slope are implausible in real survey data. This regression result reinforces the earlier finding that Study 1’s follow-up responses appear to be nearly exact copies of baseline responses.

## Q4.2

```{r q4-2}
# Study 2 control group
s2_control <- gay_r |>
  filter(study == 2, treatment == "No Contact")

fit_s2 <- lm(therm2 ~ therm1, data = s2_control)

summary(fit_s2)
```

In Study 2, the slope is 0.968662, the R-squared is 0.9476, and the residual standard error is 6.547. Compared to Study 1, the slope is meaningfully below 1, the R-squared is lower, and the residual variability is substantially larger. This pattern is consistent with real longitudinal survey data, where regression to the mean occurs and responses are not perfectly stable. In contrast, Study 1 showed a slope near 1 and an R-squared of 0.995, suggesting that therm2 was almost mechanically determined by therm1. The regression analysis therefore formalizes the earlier conclusion that Study 1’s data are implausibly perfect, while Study 2’s data appear realistic.

## Q4.3

```{r q4-3}
# Filter Study 2 to only two groups
study2_tc <- study2 |>
  filter(treatment %in% c("Same-Sex Marriage Script by Gay Canvasser",
                          "No Contact")) |>
  mutate(treated = ifelse(treatment == "Same-Sex Marriage Script by Gay Canvasser", 1, 0))

fit_group <- lm(therm2 ~ treated, data = study2_tc)

summary(fit_group)
```

The coefficient on treated is 3.5679, with a standard error of 1.2423, a t-statistic of 2.872, and a p-value of 0.00412 This matches the earlier difference in means of approximately 3.57 points, confirming that a regression with a binary treatment indicator reproduces the difference in group means. The intercept estimates the mean of the control group (57.6352). The coefficient on treated represents the average difference between the treatment and control groups. Because the p-value is below 0.05, the treatment effect is statistically significant at the 5% level.

## Q4.4

```{r q4-4}
fit_control <- lm(therm2 ~ treated + therm1, data = study2_tc)

summary(fit_control)
```

When controlling for baseline attitudes (therm1), the estimated treatment effect decreases from 3.5679 to 2.458857 points. The standard error also decreases substantially, from about 1.2423 to 0.341052, greatly improving precision. The treatment effect remains statistically significant (p < 0.001). Controlling for baseline attitudes improves precision because therm1 strongly predicts therm2. By accounting for where individuals started, the regression removes a large portion of unexplained variation in follow-up attitudes. This reduces residual variance and leads to a smaller standard error for the treatment effect.

## Q4.5

```{r q4-5}
# Predict for two hypothetical respondents
personA <- data.frame(treated = 1, therm1 = 50)
personB <- data.frame(treated = 0, therm1 = 50)

pred_A <- predict(fit_control, newdata = personA)
pred_B <- predict(fit_control, newdata = personB)

pred_A
pred_B

pred_A - pred_B
```

The R-squared increases dramatically from 0.003857 in the model without baseline controls to approximately 0.925 when therm1 is included. This indicates that baseline attitudes explain the vast majority of variation in follow-up attitudes, while the treatment explains only a small portion. Using the controlled regression model, the predicted Wave 2 score for a treated respondent with therm1 = 50 is 52.47638, while the predicted score for a control respondent with the same baseline is approximately 50.01752. The predicted difference is 2.458857 points, which matches the estimated treatment coefficient. This confirms that the coefficient on treated represents the expected difference between otherwise identical individuals who differ only in treatment status.

